### 2023-03-09 paper.tex
push minor stuff before lunch

The average thus obtained is what we use to quantify this cell's (relative) methylation in the interval.
For a genome tiled into such intervals, we thus obtain a matrix, one row per cell, one column per interval, that can be used for downstream analysis, e.g., as input for PCA.
The signal-to-noise ratio in this matrix will be better than in a matrix obtained by simply averaging absolute methylation (0 or 1) over all the cells' covered CpG sites in a region.
The reason for this is that we reduce the variation in situations as the one depicted in Figure \ref{fig:smoothres}A, where the methylation of the reads might differ strongly even though there is no actual evidence for a difference between the two cells.

@@

% remove or shorten this section cause it's not really novel, methods like this are already implemented in e.g. the "missMDA" R package
\subsection{PCA with imputation}

The standard PCA algorithm cannot deal with missing data.
Therefore, one would need to make intervals wide enough to ensure that every interval is covered by at least one read for most tiles and discard any interval that is not covered by any single cell.
Hence, a naive analysis has to resort to a coarse tiling of the genome.
To counter this, we propose a simple and straightforward way to deal with missing data in the input matrix to a PCA: In a first iteration, we replace each missing value in the centered input matrix with zero, then run the PCA.
Then, only these zeroes are replaced by the value predicted by the PCA and the PCA is rerun.
Convergence is typically achieved after two to five iterations.
For details, see Methods.

The possibility to perform PCA with missing data allows us to use shorter, and thus more intervals and hence provide the PCA with richer input data.

\subsection{Finding variably methylated regions (VMRs)}

@@

\subsection{Finding differentially methylated regions (DMRs)}

@@

Similar to the previously described VMR identification approach (Fig.~\ref{fig:vmr}), divide each chromosome into overlapping windows shifted by a small and fixed size (step size) and quantify the methylation of each cell in each window as described (Fig.~\ref{fig:smoothres}b).
Next, instead of calculating the variance, obtain the t-statistic % Welch t-test
as a measure of differential methylation between the two cell groups.
Mark the windows with the most extreme t-statistics (e.g. 2\% upper and lower tails) as DMRs.
Distinguishing between high and low t-statistics, merge any overlapping marked windows into larger DMRs.
Re-calculate the t-statistic for each larger DMR.

To assess statistical significance of DMRs, we repeat the same procedure on permutations of the scBS data.
The DMR t-statistics obtained from permuted data are then used to estimate the false-discovery rate (FDR), yielding an adjusted p-value for each DMR.

\subsection{???}  % guess we have to rewrite this closing paragraph a little bit, to include both DMRs and VMRs

Besides providing better input for the distance calculations, the small size of identified VMRs also means that they can be readily compared with genomic annotation, which aids in their functional interpretation.
While a large 100~kb tile will often contain multiple genes with distinct expression patterns, small VMRs located in known regulatory elements or gene bodies are more easily associated with individual genes and their associated function.
Thus, VMRs provide a useful starting point for examining the data with methods from functional genomics.

\section{Benchmarks} % re-name this to something more general? Application and benchmarks or something?

@@

\phantomsection
\subsection{Correlation with gene expression}

@@

\begin{figure}
    \begin{center}
    \includegraphics[width=.95\columnwidth]{figures/Fig_correlation.png}
    \end{center}
    \caption{\small \textbf{Correlation of DNA methylation and gene expression}.
    \textbf{(A)} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red).
    \textbf{(B)} Correlation of each gene's RNA expression values to methylation (mean shrunken residuals) at promoters (y-axis), or to methylation of the nearest VMR (x-axis).
    Promoters are defined as intervals \textpm2~kb around the TSS.
    \textbf{(C)} Mean methylation near the gene Htra1 (ENSMUSG00000006205).
    Cells are assigned to 4 groups based on Htra1 expression (group 0: cells with no Htra1 expression, groups 1-3: cells that express Htra1, divided into three equally large groups with group 3 having the highest expression).
    Data from \citet{kremer_scnmt}.}
    \label{fig:correlation}
\end{figure}

>>>>

The average thus obtained, the shrunken mean of the residuals, is what we use to quantify this cell's (relative) methylation in the interval.
For a genome tiled into such intervals, we thus obtain a matrix, one row per cell, one column per interval, that can be used for downstream analysis, e.g., as input for PCA.
The signal-to-noise ratio in this matrix will be better than in a matrix obtained by simply averaging absolute methylation (0 or 1) over all the cells' covered CpG sites in a region.
The reason for this is that we reduce the variation in situations as the one depicted in Figure \ref{fig:smoothres}A, where the methylation of the reads might differ strongly even though there is no actual evidence for a difference between the two cells.

@@

% remove or shorten this section cause it's not really novel, methods like this are already implemented in e.g. the "missMDA" R package
%\subsection{PCA with imputation}
%The standard PCA algorithm cannot deal with missing data.
%Therefore, one would need to make intervals wide enough to ensure that every interval is covered by at least one read for most tiles and discard any interval that is not covered by any single cell.
%Hence, a naive analysis has to resort to a coarse tiling of the genome.
%To counter this, we propose a simple and straightforward way to deal with missing data in the input matrix to a PCA: In a first iteration, we replace each missing value in the centered input matrix with zero, then run the PCA.
%Then, only these zeroes are replaced by the value predicted by the PCA and the PCA is rerun.
%Convergence is typically achieved after two to five iterations.
%For details, see Methods.
%The possibility to perform PCA with missing data allows us to use shorter, and thus more intervals and hence provide the PCA with richer input data.

@@

\subsection{Finding differentially methylated regions}

@@

Similar to the previously described VMR identification approach (Fig.~\ref{fig:vmr}), divide each chromosome into overlapping windows shifted by a small and fixed size (step size) and quantify the methylation of each cell in each window.
Next, instead of the variance, obtain the t-statistic % Welch t-test
as a measure of differential methylation between the two cell groups.
Identify the windows with the most extreme t-statistics, e.g. windows in the 2\% upper and lower tails.
Merge any overlapping windows in the upper tail into larger DMRs.
Repeat for windows in the lower tail, then re-calculate the t-statistic for each larger DMR.

To assess statistical significance of DMRs, repeat the same procedure on permutations of the scBS data, i.e. the same data set with randomly shuffled cell labels.
The DMR t-statistics obtained from permuted data are then used to estimate the false-discovery rate (FDR), yielding an adjusted p-value for each DMR.

\subsection{Utility of DMRs and VMRs} % not sure if we need a subsection here

While the primary purpose of VMRs is to provide better input for the distance calculations, their small size also means that they can be readily compared with genomic annotation, which aids in their functional interpretation.
Large 100~kb tiles will often contain multiple genes with distinct expression patterns, small VMRs located in known regulatory elements or gene bodies are more easily associated with individual genes and their associated function.
Thus, VMRs provide a useful starting point for examining the data with methods from functional genomics.
To enable a more targeted characterization of 
The same principles may also be applied to DMRs, thus enabling in-depth characterization of epigenomic and gene regulatory differences between tissues, cell types or conditions.

\section{Application and benchmarks}

@@

\phantomsection
\subsection{Correlation of VMR methylation with gene expression}

The first data set used to test our methods is a single-cell multi-omics data set generated by \citet{kremer_scnmt}.
This data set comprises the methylomes of 1566 cells isolated from mouse forebrains, as well as single-cell transcriptomes of the same cells.

@@

\begin{figure}
    \begin{center}
    \includegraphics[width=.95\columnwidth]{figures/Fig_correlation.png}
    \end{center}
    \caption{\small \textbf{Correlation of DNA methylation and gene expression}.
    \textbf{(A)} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red).
    \textbf{(B)} Correlation of each gene's RNA expression values to methylation (shrunken mean of residuals) at promoters (y-axis), or to methylation of the nearest VMR (x-axis).
    Promoters are defined as intervals \textpm2~kb around the TSS.
    \textbf{(C)} Mean methylation near the gene Htra1 (ENSMUSG00000006205).
    Cells are assigned to 4 groups based on Htra1 expression (group 0: cells with no Htra1 expression, groups 1-3: cells that express Htra1, divided into three equally large groups with group 3 having the highest expression).
    Data from \citet{kremer_scnmt}.}
    \label{fig:correlation}
\end{figure}
